Reformulating phylogenetic mixed models to improve flexibility and speed

### 2018-09-11 main.tex
adding simulation full model formulas

\begin{equation}
\bZ^{C}_{i} = (\bS^{T}_{adji}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T}, \label{eq:ZC}
\end{equation}
where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}, $\bS_{adji}$ is an $m_{i} \times b_{i}$ matrix of the scaled species--branch matrix; $\bJ_{i}$ the indicator matrix of grouping factors indices matrix size $n \times m_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.

@@

The scaled species--branch matrix $\bS_{adji}$ is scaled version of $\bS$ via the standardized generalized variance (SGV).
SGVs are used as an intrinsic/natural way to scale down the variance-covariance matrix measured in different scales (for example, phylogenetic distance, years and etc) compariable to other variation in the model \citep{sengupta1987tests}.
SGV is the $nth$ root of the generalized variance (GV), or the determinant of the variance-covariance matrix, where GV measures the overall size or volume of the variance-covariance matrix.  
The SGV is given by:
\begin{align}
\bomega & = \textrm{det}(\bSigma_{phy})^{1/n},
\end{align}
where $n$ is the number of species in the phylogeny.

The analog in the $\bS$ is to scale the branch vector $\bell$ by square root of the SGV:
\begin{align}
\bell_{adj} & = \frac{\bell}{\sqrt{\bomega}}
\end{align} 
and the $\bS_{adj}$ is the matrix product of $\bS_{ind}$ and $\bell_{adj}$.

@@

To account for multiple levels of phylogenetic signals, we assume phylogenetic signals in the intercept $b_0$, and slope $b_1$. \bmb{clarify what you mean by ``multiple levels of phylog. signals'' here?}
We further assume that random slope and intercept are correlated. \bmb{, ``i.e. ...'' it will help readers unfamiliar with these ideas to give slightly more detailed examples}
For models with multiple sites, we also assume for species and site interactions random effects, an equivalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.

>>>>

\begin{equation}
\bZ^{C}_{i} = (\bS^{T}_{i}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T}, \label{eq:ZC}
\end{equation}
where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}, $\bS_{i}$ is an $m_{i} \times b_{i}$ matrix of the species--branch matrix; $\bJ_{i}$ the indicator matrix of grouping factors indices matrix size $n \times m_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.

@@

% 
% The scaled species--branch matrix $\bS_{adji}$ is scaled version of $\bS$ via the standardized generalized variance (SGV).
% SGVs are used as an intrinsic/natural way to scale down the variance-covariance matrix measured in different scales (for example, phylogenetic distance, years and etc) compariable to other variation in the model \citep{sengupta1987tests}.
% SGV is the $nth$ root of the generalized variance (GV), or the determinant of the variance-covariance matrix, where GV measures the overall size or volume of the variance-covariance matrix.  
% The SGV is given by:
% \begin{align}
% \bomega & = \textrm{det}(\bSigma_{phy})^{1/n},
% \end{align}
% where $n$ is the number of species in the phylogeny.
% 
% The analog in the $\bS$ is to scale the branch vector $\bell$ by square root of the SGV:
% \begin{align}
% \bell_{adj} & = \frac{\bell}{\sqrt{\bomega}}
% \end{align} 
% and the $\bS_{adj}$ is the matrix product of $\bS_{ind}$ and $\bell_{adj}$.

@@

The full single site simulation model contains two fixed effects parameters ($\beta_0$ and $\beta_1$), and four random effect parameters (phylogenetic random intercept, phylogenetic random slope, correlation between phylogenetic random slope and intercept and observation variance).  
Predictor and intercept level random variation/effect of species are not applicable in this simulation setting with single observation per species because it will be confounded with the observational level variation (residuals).

The full multi-site simulation model :

\begin{align}
\bY & = (\beta_0 + b_{phy_{int}} + b_{sp_{int}} + b_{site}) + (\beta_1 + b_{phy_{slope}} + b_{sp_{slope}}) X + b_{sp:site interaction} + b_{obs}\\
(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
\sigma^2_{phy_{int}} & \sigma_{phy_{int},phy_{slope}} \\ 
\sigma_{phy_{int},phy_{slope}} & \sigma^2_{phy_{slope}}
\end{bmatrix}
\Bigg) \\
(b_{sp_{int}}, b_{sp_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
\sigma^2_{sp_{int}} & \sigma_{sp_{int},phy_{slope}} \\ 
\sigma_{sp_{int},sp_{slope}} & \sigma^2_{sp_{slope}}
\end{bmatrix}
\Bigg) \\
b_{site} & \sim \textrm{MVN} ( 0 , \Sigma_{site}) \\
b_{sp:site} & \sim \textrm{MVN} (0, kron(\bf{I}_{site}, \sigma^2_{phy})) \\
b_{obs} & \sim \textrm{MVN} ( 0 , \Sigma)
\end{align}

% 
% To account for multiple levels of phylogenetic signals, we assume phylogenetic signals in the intercept $b_0$, and slope $b_1$. \bmb{clarify what you mean by ``multiple levels of phylog. signals'' here?}
% We further assume that random slope and intercept are correlated.

\bmb{, ``i.e. ...'' it will help readers unfamiliar with these ideas to give slightly more detailed examples}
For models with multiple sites, we also assume for species and site interactions random effects, an equivalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.
